---
title: "Reproducible Research: Peer Assessment 2, Storms"
output: 
  html_document:
    keep_md: true
---


## Loading packages
```{r}
library(dplyr)
library(lubridate)
```


### Loading and Processing the Raw data

We download the data from the course website for Reproducible Research. Note that https:// URLs are not supported by download.file by default, but using method = "curl" permits the download. Also we use "bzfile" in the read.csv argument since the file is a bz2 file.

```{r, cache=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp, method = "curl")
data <- read.csv(bzfile(temp, "repdata%2Fdata%2FStormData.csv.bz2"), header = TRUE)
unlink(temp)
```

Let's have a look at the data. First of all I prefer having variables in lowercase
```{r}
data2 <- data ## remove this later, just so I don't have to keep reading in the file
data2 <- data2[,c("bgn_date", "state", "evtype",  "fatalities", "injuries", "propdmg", "propdmgexp", "cropdmg", "cropdmgexp")]

names(data2) <- tolower(names(data2))
names(data2)
data2$evtype <- tolower(data2$evtype)
length(unique(data2$evtype))
```


```{r, cache=TRUE}
## remove extraneous variables


head(data2)

## remove extraneous event types - working
event_pattern <- "(summary|\\?|other)"
data2$index <- grepl(event_pattern, data2$evtype)
data2 <- data2[!data2$index == TRUE,] # remove daily summaries and unknown types


## group thunderstorm types together - working
length(unique(data2$evtype))
data2$evtype2 <- data2$evtype
length(unique(data2$evtype2))
thunder_pattern <- "(thunder|microburst)"
data2$evtype2 <- ifelse(grepl(thunder_pattern, data2$evtype) == TRUE, "thunderstorm", data2$evtype2)

length(unique(data2$evtype2))
```


# Question 1:
## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?




# Question 2:
## Across the United States, which types of events have the greatest economic consequences?

