---
title: "Reproducible Research: Peer Assessment 2, Storms"
output: 
  html_document:
    keep_md: true
---


## Loading packages
```{r}
library(dplyr)
library(lubridate)
```


### Loading and Processing the Raw data

We download the data from the course website for Reproducible Research. Note that https:// URLs are not supported by download.file by default, but using method = "curl" permits the download. Also we use "bzfile" in the read.csv argument since the file is a bz2 file.

```{r, cache=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp, method = "curl")
data <- read.csv(bzfile(temp, "repdata%2Fdata%2FStormData.csv.bz2"), header = TRUE)
unlink(temp)
```

Let's have a look at the data. First of all I prefer having variables in lowercase
```{r}
data2 <- data ## remove this later, just so I don't have to keep reading in the file
names(data2) <- tolower(names(data2))
data2 <- data2[,c("bgn_date", "state", "evtype",  "fatalities", "injuries", "propdmg", "propdmgexp", "cropdmg", "cropdmgexp")]


names(data2)
data2$evtype <- tolower(data2$evtype)
length(unique(data2$evtype))
```


```{r, cache=TRUE}
## remove extraneous variables


head(data2)

## remove extraneous event types - working
event_pattern <- "(summary|\\?|other|none|month|no severe)"
data2$index <- grepl(event_pattern, data2$evtype)
data2 <- data2[!data2$index == TRUE,] # remove daily summaries and unknown types


## group thunderstorm types together - working
length(unique(data2$evtype))
data2$evtype <- gsub(" ", "_", data2$evtype)
data2$evtype2 <- data2$evtype
length(unique(data2$evtype2))
thunder_pattern <- "(thund|microburst|tstm|tunder|thuder)"
data2$evtype2 <- ifelse(grepl(thunder_pattern, data2$evtype) == TRUE, "thunderstorm", data2$evtype2)

length(unique(data2$evtype2))

hurricane_pattern <- "(hurricane|typhoon)"
data2$evtype2 <- ifelse(grepl(hurricane_pattern, data2$evtype) == TRUE, "hurricane", data2$evtype2)

tropical_storm_pattern <- "tropical storm"
data2$evtype2 <- ifelse(grepl(tropical_storm_pattern, data2$evtype) == TRUE, "tropical storm", data2$evtype2)

drought_pattern <- "(dry|driest|record_low_rainfall)"
data2$evtype2 <- ifelse(grepl(drought_pattern, data2$evtype) == TRUE, "drought", data2$evtype2)

cold_pattern <- "(cold|wind_chill|windchill|record_low|low_temp)"
data2$evtype2 <- ifelse(grepl(cold_pattern, data2$evtype) == TRUE, "cold_windchill", data2$evtype2)

heat_pattern <- "(heat|hot|warm|record_temperature|hyperthermia|temperature_record|record_high)"
data2$evtype2 <- ifelse(grepl(heat_pattern, data2$evtype) == TRUE, "heat", data2$evtype2)

surf_pattern <- "surf"
data2$evtype2 <- ifelse(grepl(surf_pattern, data2$evtype) == TRUE, "high_surf", data2$evtype2)

tornado_pattern <- "nado|torn"
data2$evtype2 <- ifelse(grepl(tornado_pattern, data2$evtype) == TRUE, "tornado", data2$evtype2)

hail_pattern <- "(hail|ice_pellets)"
data2$evtype2 <- ifelse(grepl(hail_pattern, data2$evtype) == TRUE, "hail", data2$evtype2)

flood_pattern <- "(flood|fld)"
data2$evtype2 <- ifelse(grepl(flood_pattern, data2$evtype) == TRUE, "flood", data2$evtype2)

rain_pattern <- "(rain|record_precipitation|wet)"
data2$evtype2 <- ifelse(grepl(rain_pattern, data2$evtype) == TRUE, "heavy rain", data2$evtype2)

avalanche_pattern <- "avalan"
data2$evtype2 <- ifelse(grepl(avalanche_pattern, data2$evtype) == TRUE, "avalanche", data2$evtype2)

blizzard_pattern <- "blizz"
data2$evtype2 <- ifelse(grepl(blizzard_pattern, data2$evtype) == TRUE, "blizzard", data2$evtype2)


fire_pattern <- "fire"
data2$evtype2 <- ifelse(grepl(fire_pattern, data2$evtype) == TRUE, "fire", data2$evtype2)

winter_pattern <- "winter"
data2$evtype2 <- ifelse(grepl(winter_pattern, data2$evtype) == TRUE, "winter weather", data2$evtype2)

frost_pattern <- "(frost|freez)"
data2$evtype2 <- ifelse(grepl(frost_pattern, data2$evtype) == TRUE, "frost_freeze", data2$evtype2)

sleet_pattern <- "sleet"
data2$evtype2 <- ifelse(grepl(sleet_pattern, data2$evtype) == TRUE, "sleet", data2$evtype2)

fog_pattern <- "fog"
data2$evtype2 <- ifelse(grepl(fog_pattern, data2$evtype) == TRUE, "fog", data2$evtype2)

lightning_pattern <- "lightning"
data2$evtype2 <- ifelse(grepl(lightning_pattern, data2$evtype) == TRUE, "lightning", data2$evtype2)

snow_ice_pattern <- "(snow|ice)" # note many events list both so hard to distinguish
data2$evtype2 <- ifelse(grepl(snow_ice_pattern, data2$evtype) == TRUE, "snow_ice", data2$evtype2)

tide_pattern <- "(high_tide|surge)"
data2$evtype2 <- ifelse(grepl(tide_pattern, data2$evtype) == TRUE, "tide_surge", data2$evtype2)

length(unique(data2$evtype2))

data2 %>% group_by(evtype2) %>% summarise(fatalities = sum(fatalities)) %>% arrange(desc(fatalities))
```


# Question 1:
## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?




# Question 2:
## Across the United States, which types of events have the greatest economic consequences?

